---
title: "Making R Notebooks Prettier"
author: "Cynthia Deng"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
if (!require(ggthemes)) install.packages("ggthemes")
library(ggthemes)
```

---

# Introduction

Goal: after completing a series of portfolios, I wanted to step back
and explore how to make them prettier.

The dataset used throughout is the built-in `iris` dataset. It has 150 flowers
across 3 species (*setosa*, *versicolor*, *virginica*), with measurements for
sepal and petal length and width.

---

# Section 1: YAML Header Options

The YAML header is the block at the very top of your `.Rmd` file, sitting between
two sets of `---` lines. It tells R how to knit the document. Here is what this
notebook's header looks like:

```yaml
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
```

Each line is an option. Below is what every single one does.

## toc

`toc` stands for Table of Contents.

```yaml
toc: true
```

Setting this to `true` adds a navigation panel to your document that lists all
your section headings. Without it, there is no navigation and readers have to
scroll through the whole page to find what they want.

Setting it to `false` (or just leaving it out) removes the TOC entirely.

```yaml
toc: false   # no table of contents at all
```

## toc_float

```yaml
toc_float: true
```

By default, even if you turn the TOC on, it sits at the top of the page and
disappears as soon as you scroll down. Setting `toc_float: true` makes it stick
to the left side of the screen so it is always visible no matter how far down
the page you are.

Setting it to `false` puts the TOC back at the top and makes it static.

```yaml
toc_float: false   # TOC sits at the top and scrolls away with the page
```

You can also pass extra options to `toc_float` to control its behaviour:

```yaml
toc_float:
  collapsed: false   # show all sub-sections expanded by default
  smooth_scroll: false   # jump to sections instead of smooth scrolling
```

By default `collapsed` is `true`, meaning only top-level headings show until
you click them. Setting it to `false` shows everything expanded from the start.

## toc_depth

```yaml
toc_depth: 3
```

This controls how many levels of headings appear in the TOC.

In Markdown, headings are made with `#` symbols. One `#` is the biggest heading,
`##` is one level smaller, `###` is smaller still, and so on.

`toc_depth: 3` means headings at levels `#`, `##`, and `###` all show up in the
TOC. If you change it to `2`, only `#` and `##` appear, and `###` headings are
ignored in the navigation panel (they still appear in the document, just not in
the TOC).

```yaml
toc_depth: 2   # only top two heading levels appear in the TOC
toc_depth: 4   # four levels deep, including ####
```

## theme

```yaml
theme: flatly
```

This sets the overall look of the whole HTML page, including fonts, colors,
button styles, and spacing. The themes come from a free library called
[Bootswatch](https://bootswatch.com/). Changing this one word completely
transforms the look of the document.

**Try swapping `flatly` for any of these and re-knitting to see the difference:**

| Theme | What it looks like |
|-------|--------------------|
| `default` | Plain Bootstrap, no frills |
| `flatly` | Clean, flat, modern. A good all-purpose default |
| `cerulean` | Blue accents, professional |
| `journal` | Serif fonts, feels like a magazine |
| `readable` | Slightly larger fonts, easy on the eyes |
| `spacelab` | Light and airy with lots of white space |
| `united` | Orange accents, inspired by Ubuntu |
| `cosmo` | Clean and rounded, modern feel |
| `sandstone` | Warm earthy tones |
| `yeti` | Bold headings, very clean |
| `darkly` | Dark mode |
| `cyborg` | Dark background with blue accents |
| `slate` | Dark grey tones |
| `superhero` | Dark background with orange accents |

> 💡 To try a theme, change `theme: flatly` in the YAML header to any name from
> the table above, then click Knit. The entire page updates instantly.

## highlight

```yaml
highlight: tango
```

This only affects one thing: the colours used inside code chunks. It controls
how keywords, strings, numbers, and comments are coloured in your code blocks.
It has no effect on the rest of the page.

**Available highlight styles:**

| Style | Feel |
|-------|------|
| `default` | Simple, muted colours |
| `tango` | Colourful and easy to read. Very popular |
| `pygments` | Inspired by the Python syntax highlighter |
| `kate` | From the Kate text editor |
| `monochrome` | Black and white only, no colours |
| `espresso` | Dark code background |
| `zenburn` | Soft dark theme, easy on the eyes |
| `haddock` | Subtle and muted |
| `breezedark` | Modern dark style |
| `textmate` | From the TextMate editor |

> 💡 `highlight` only changes how code looks inside backtick blocks.
> It does not affect headings, body text, or anything else on the page.

---

# Section 2: ggthemes

The `ggthemes` package adds extra themes to ggplot2. These are themes inspired
by real publications and tools. Once the package is loaded with
`library(ggthemes)`, the themes are available just like `theme_minimal()` or
`theme_classic()`.

The same base plot is used throughout so the differences are easy to compare.
Here is that base plot with no theme applied yet:

```{r base_plot}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 2.5, alpha = 0.8) +
  labs(
    title = "Sepal Length vs Width",
    x     = "Sepal Length (cm)",
    y     = "Sepal Width (cm)"
  )
```

This is just the default ggplot2 style. Now the same plot is run through three
different `ggthemes` themes below.

## theme_economist

This theme mimics the style used by *The Economist* magazine. It has a blue-grey
background, white grid lines, and a distinctive look that feels very editorial.

```{r economist_basic}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 2.5, alpha = 0.8) +
  labs(
    title    = "Sepal Length vs Width",
    subtitle = "theme_economist()",
    x        = "Sepal Length (cm)",
    y        = "Sepal Width (cm)"
  ) +
  theme_economist()
```

The theme changes the background, fonts, and grid but leaves the default ggplot2
colour palette on the points. To match the colours to the theme as well, add
`scale_color_economist()`:

```{r economist_with_scale}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 2.5, alpha = 0.8) +
  labs(
    title    = "Sepal Length vs Width",
    subtitle = "theme_economist() + scale_color_economist()",
    x        = "Sepal Length (cm)",
    y        = "Sepal Width (cm)"
  ) +
  theme_economist() +
  scale_color_economist()
```

`scale_color_economist()` replaces the default ggplot2 colours with the colour
palette used by *The Economist*. Pairing a theme with its matching scale makes
the whole plot feel consistent.

## theme_fivethirtyeight

This theme is inspired by the data journalism site FiveThirtyEight. It has a
grey background, no axis titles by default, and a clean editorial feel.

```{r fte_basic}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 2.5, alpha = 0.8) +
  labs(
    title    = "Sepal Length vs Width",
    subtitle = "theme_fivethirtyeight()",
    x        = "Sepal Length (cm)",
    y        = "Sepal Width (cm)"
  ) +
  theme_fivethirtyeight()
```

Notice that the axis labels (`x` and `y`) from `labs()` do not appear. This is
intentional. The FiveThirtyEight style drops axis titles. To bring them back,
add `theme(axis.title = element_text())` after the theme call:

```{r fte_with_axis}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 2.5, alpha = 0.8) +
  labs(
    title    = "Sepal Length vs Width",
    subtitle = "theme_fivethirtyeight() with axis titles restored",
    x        = "Sepal Length (cm)",
    y        = "Sepal Width (cm)"
  ) +
  theme_fivethirtyeight() +
  theme(axis.title = element_text())
```

Adding `theme()` after `theme_fivethirtyeight()` lets you override individual
parts of the theme without replacing the whole thing. Here it turns axis titles
back on. Now with the matching colour scale added too:

```{r fte_full}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 2.5, alpha = 0.8) +
  labs(
    title    = "Sepal Length vs Width",
    subtitle = "theme_fivethirtyeight() + scale_color_fivethirtyeight()",
    x        = "Sepal Length (cm)",
    y        = "Sepal Width (cm)"
  ) +
  theme_fivethirtyeight() +
  scale_color_fivethirtyeight() +
  theme(axis.title = element_text())
```

## theme_tufte

This theme is based on the design principles of Edward Tufte, a statistician
known for writing about data visualisation. His philosophy is to remove anything
that does not add information. The result is a very stripped-back plot with no
background, no grid lines, and almost no decoration at all.

```{r tufte_basic}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 2.5, alpha = 0.8) +
  labs(
    title    = "Sepal Length vs Width",
    subtitle = "theme_tufte()",
    x        = "Sepal Length (cm)",
    y        = "Sepal Width (cm)"
  ) +
  theme_tufte()
```

`theme_tufte()` removes grid lines and the axis box entirely, leaving only the
axis lines themselves. It can look very plain at first but it puts all the
attention on the data points.

A common pairing with `theme_tufte()` is `geom_rangeframe()`, which draws short
axis lines only over the range of the data instead of running the full length:

```{r tufte_rangeframe}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 2.5, alpha = 0.8) +
  geom_rangeframe() +
  labs(
    title    = "Sepal Length vs Width",
    subtitle = "theme_tufte() + geom_rangeframe()",
    x        = "Sepal Length (cm)",
    y        = "Sepal Width (cm)"
  ) +
  theme_tufte()
```

`geom_rangeframe()` comes from `ggthemes` as well. It adds short tick-style axis
lines that only span the range of the actual data. The combination of
`theme_tufte()` and `geom_rangeframe()` is very common when using this theme.

---

# Section 3: Custom CSS

Everything covered so far changes the look of plots or the page theme. CSS goes
one step further and lets you control the look of the page itself directly.

CSS stands for Cascading Style Sheets. It is the language that browsers use to
decide how things look on a webpage. Since R Markdown knits to HTML, any CSS
rules you write get applied to the final document.

You do not need to be a web developer to use CSS in a notebook. A few simple
rules can make a big visual difference.

## How to add CSS to your notebook

You add a chunk but instead of `{r}` you write `{css}`:

````markdown
```{css}
/* your rules go here */
```
````

That is all. When you knit, those rules apply to the whole page. Everything
inside `/* */` is a comment and does not do anything, it is just a note to
yourself.

Every CSS rule follows the same pattern:

```
selector {
  property: value;
}
```

The **selector** is what you are targeting (for example `h1` targets all level-1
headings). The **property** is what you want to change (for example `color`).
The **value** is what to change it to (for example `red` or `#2c7bb6`).

## Styling 1: Headings

This rule adds a coloured bottom border to all `h1` headings and changes the
colour of all `h2` headings.

```{css}
/* Style all h1 headings */
h1 {
  color: #2c3e50;
  border-bottom: 3px solid #2c7bb6;
  padding-bottom: 6px;
}

/* Style all h2 headings */
h2 {
  color: #2c7bb6;
}
```

What each property does:

- `color` sets the text colour. You can use a colour name like `red` or a hex
  code like `#2c7bb6`. Hex codes give you more precise control.
- `border-bottom` draws a line under the element. The three values are the
  thickness (`3px`), the line style (`solid`), and the colour (`#2c7bb6`).
  You could also write `dashed` or `dotted` instead of `solid`.
- `padding-bottom` adds space between the text and the line below it. Without
  this the border sits right against the text.

Try changing `#2c7bb6` to another hex code or colour name to see a different
result. For example `#e74c3c` is red, `#27ae60` is green, `#8e44ad` is purple.

## Styling 2: Body font

This rule changes the font size and line spacing for the whole page.

```{css}
/* Change the font size and line spacing for the whole page */
body {
  font-size: 15px;
  line-height: 1.7;
}
```

What each property does:

- `font-size` sets how big the text is. The default in most themes is around
  `14px`. Bumping it to `15px` or `16px` makes the page easier to read.
- `line-height` controls the spacing between lines. A value of `1.7` means each
  line is 1.7 times the font size tall, which adds breathing room between lines.
  The default is usually around `1.5`.

Try changing `font-size` to `18px` and `line-height` to `2.0` to see how
spacious the page feels, then bring it back down.

## Styling 3: Callout boxes

A callout box is a styled block you can use to highlight tips or important notes.
CSS does not create the box automatically, you have to attach a class to a
`<div>` tag in your markdown.

First, define the class in CSS:

```{css}
/* A blue note box */
.note {
  background-color: #f0f7ff;
  border-left: 5px solid #2c7bb6;
  padding: 12px 16px;
  margin: 16px 0;
  border-radius: 4px;
}

/* A red warning box */
.warning {
  background-color: #fff5f5;
  border-left: 5px solid #e74c3c;
  padding: 12px 16px;
  margin: 16px 0;
  border-radius: 4px;
}
```

What each property does:

- `background-color` fills the inside of the box with a colour.
- `border-left` draws a thick coloured stripe on the left edge. The three
  values are thickness, style, and colour, same as `border-bottom` above.
- `padding` adds space between the text and the edges of the box. Writing
  `12px 16px` is shorthand for 12px top and bottom, 16px left and right.
- `margin` adds space outside the box, between the box and whatever is above
  or below it. `16px 0` means 16px top and bottom, 0 left and right.
- `border-radius` rounds the corners. A value of `4px` is a subtle rounding.
  Try `12px` for rounder corners or `0` for sharp square corners.

Then use the class in your markdown like this:

```html
<div class="note">
💡 This is a note box.
</div>

<div class="warning">
⚠️ This is a warning box.
</div>
```

Which produces these:

<div class="note">
💡 This is a note box. Use it to highlight tips or anything you want the reader
to pay attention to.
</div>

<div class="warning">
⚠️ This is a warning box. Same structure as the note box, just different colours.
</div>

## Styling 4: Code chunks

You can also style the appearance of code chunks on the page. In the knitted
HTML, code chunks are wrapped in a `<pre>` tag. Targeting `pre` in CSS lets you
change how they look.

```{css}
/* Style code chunk blocks */
pre {
  background-color: #f8f8f8;
  border: 1px solid #e0e0e0;
  border-radius: 6px;
  padding: 12px;
  font-size: 13px;
}
```

What each property does:

- `background-color` sets the background of the code block. `#f8f8f8` is a
  very light grey, which is softer than bright white.
- `border` draws a line around the entire block. The three values are thickness,
  style, and colour. `1px solid #e0e0e0` gives a subtle light grey border.
- `border-radius` rounds the corners of the block, same as the callout boxes.
- `padding` adds space between the code text and the edges of the block.
- `font-size` controls the size of the text inside code blocks specifically,
  without affecting the rest of the page.

---

# Summary

| Section | Tool | What was explored |
|---------|------|-------------------|
| YAML | `toc`, `toc_float`, `toc_depth` | Navigation panel options |
| YAML | `theme` | Full page visual themes from Bootswatch |
| YAML | `highlight` | Code syntax colour styles |
| ggthemes | `theme_economist()` | Editorial style with blue-grey background |
| ggthemes | `theme_fivethirtyeight()` | Data journalism style, grey background |
| ggthemes | `theme_tufte()` | Minimal, data-first, no decoration |
| ggthemes | `scale_color_*()` | Matching colour palettes for each theme |
| CSS | headings | Colour and border styling for `h1` and `h2` |
| CSS | body | Font size and line height for the whole page |
| CSS | callout boxes | Custom styled blocks using `<div class="...">` |
| CSS | code chunks | Background, border, and font size for `pre` blocks |

---
